Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Tleaf section edits

![**Figure `r fig_leaf_T`. Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`. In each scenario, an the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants. swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (cm); gs = stomatal conductance (mmol m^2^/s). ](leaf_energy_balance/fig3_leafenergy.png)

@@

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).
Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
In contrast to closed canopies, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy. 
The latter can result from the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].

In forests with dense canopies, including tropical and temperate broadleaf forests, leaves in the upper canopy experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].

@@

Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].

Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].

>>>>

![**Figure `r fig_leaf_T`. Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`. In each scenario, an the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants. swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (m); gs = stomatal conductance (mmol m^2^/s). ](leaf_energy_balance/fig3_leafenergy.png)

@@

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b]. Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].

@@

Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. 
Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
